D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-04-09 01intro_preliminary.tex
.

\subsubsection*{Contributions}
This paper presents D-Hammer, an automated tool for reasoning about
labelled Dirac notation. D-Hammer uses a rich, dependently typed
language to formalize labelled Dirac notation and supports common
idioms for describing quantum systems, including big operators of the
form \( \sum_{i \in I} a_i \ket{\phi_i}_S \) to represent indexed
superpositions of states. The semantics of typable expressions are
given in terms of Hilbert spaces, tailored to interpret the tensor
product as an AC symbol. We leverage this interpretation to define a
rich equational theory for labelled Dirac notation and prove its
soundness with respect to our denotational semantics. Finally, we
define an efficient normalization procedure to prove equivalence
between two expressions. We then evaluate our procedure with respect
to examples from the literature. Our evaluation covers examples in
plain and labelled Dirac notation. The main conclusions are that our
approach outperforms DiracDec~\cite{diracdec} to reason about plain
Dirac notation and is able of proving complex examples from the
literature on labelled Dirac notation, including examples from prior
work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
For completeness, we also evaluate D-Hammer on examples from the
literature on equivalence checking of (parametrized) quantum
circuits.\gb{want to say our tool is not competitive?}

>>>>

\subsubsection*{Contributions}
This paper presents D-Hammer, an automated tool for reasoning about
labelled Dirac notation. D-Hammer uses a rich, dependently typed
language to formalize labelled Dirac notation and supports common
idioms for describing quantum systems, including big operators of the
form \( \sum_{i \in I} a_i \ket{\phi_i}_S \) to represent indexed
superpositions of states. The semantics of typable expressions are
given in terms of Hilbert spaces, tailored to interpret the tensor
product as an AC symbol. We leverage this interpretation to define a
rich equational theory for labelled Dirac notation and prove its
soundness with respect to our denotational semantics. Finally, we
define an efficient normalization procedure to prove equivalence
between two expressions. We then evaluate our procedure with respect
to examples from the literature. Our evaluation covers examples in
plain and labelled Dirac notation. The main conclusions are that our
approach outperforms DiracDec~\cite{diracdec} to reason about plain
Dirac notation and is able of proving complex examples from the
literature on labelled Dirac notation, including examples from prior
work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
For completeness, we also evaluate D-Hammer on examples from the
literature on equivalence checking of (parametrized) quantum
circuits.

### 2025-04-09 03labelled.tex
.

\paragraph*{Typing rules.}
There are various rules for computing types and checking vadility of
registers and labelled terms. Here we display some of the rules and
refer the reader to Appendix for the full set of rules.\gb{which appendix? add section}
\begin{gather*}
  \frac{
      \Gamma \vdash R : \reg(\sigma) \quad
      \Gamma \vdash Q : \reg(\tau)
      \quad \var(R) \cap \var(Q) = \emptyset
  }{\Gamma \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
  \frac{
          \Gamma \vdash r : \reg(\sigma) \quad
          \Gamma \vdash i : \Basis(\sigma)}
  {\Gamma \vdash |i\>_r : \DType(\{r\}, \emptyset)}
  \qquad
  \frac{\Gamma \vdash R : \reg(\sigma)\qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
    \frac{
        \Gamma \vdash D_i : \DType(s_i,s_i') \qquad
        \forall\,i\neq j.\ s_i\cap s_j = \emptyset \qquad
        \forall\,i\neq j.\ s'_i\cap s'_j = \emptyset
    }
    {\Gamma \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\end{gather*}
The first rule states that a paired register is of the product type and its components must be disjoint.
To lift a plain Dirac notation into the labelled version (line 2), we enforce that the term and register share the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
The third line provides the typing of labelled tensor product, with a check to ensure that the component subsystems are disjoint from each other.

@@

The procedure to eliminate labels is sound and complete, in the sense
that expressions in Labelled Dirac notation are equivalent iff their
translations to plain Dirac notation are equivalent.
\begin{theorem}[Label Elimination]
  \label{thm: normalization}
  Assume $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 :
  \cD(s,s')$ and no variables of $\DType(\cdot,\cdot)$ appear in
  $D_1,D_2$.  Let $e_1 = e_2$ be obtained by above normalization
  procedure on $D_1 = D_2$.  Then $e!=e_2$ is an equation in plain
  Dirac notation and $\Gamma\vDash D_1 = D_2$ if and only if
  $\Gamma\vDash e_1 = e_2$.
\end{theorem}
The idea of the proof is as follows: first, we define a set of proof
rules to rewrite every labelled Dirac notation into the form of
Eqn. (1). We prove that each rule is sound w.r.t.\, semantics, and
that every labelled Dirac expression can be rewritten to an expression
of the form of Eqn. (1); the proof is by induction on the structure of
the expression. Next, we show that reordering of labelled basis
preserves the type and semantics and thus yield expressions of the
form of Eqn. (4), as desired. Details appear in
Appendix~\ref{app:label_elim}.\gb{check proof is really there}
%% \begin{hproof}
%%   \lz{
%%   Checking the soundness of every rule is routine. 
%%   For relative completeness, it is sufficient to show that the normalization steps can succeed for every labelled Dirac equality. To this end, we give 
%%   concrete steps to  by induction on the structure of the notation. Further noticing that reordering of labelled basis preserves the type and semantics and thus yield Eqn. (4). The detailed proof can be found in Appendix \ref{sec: rewriting rules}.
%%   }

>>>>

\paragraph*{Typing rules.}
There are various rules for computing types and checking vadility of
registers and labelled terms. Here we display some of the rules and
refer the reader to~\Cref{sec: full typing rules} for the full set of rules.
\begin{gather*}
  \frac{
      \Gamma \vdash R : \reg(\sigma) \quad
      \Gamma \vdash Q : \reg(\tau)
      \quad \var(R) \cap \var(Q) = \emptyset
  }{\Gamma \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
  \frac{
          \Gamma \vdash r : \reg(\sigma) \quad
          \Gamma \vdash i : \Basis(\sigma)}
  {\Gamma \vdash |i\>_r : \DType(\{r\}, \emptyset)}
  \qquad
  \frac{\Gamma \vdash R : \reg(\sigma)\qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
    \frac{
        \Gamma \vdash D_i : \DType(s_i,s_i') \qquad
        \forall\,i\neq j.\ s_i\cap s_j = \emptyset \qquad
        \forall\,i\neq j.\ s'_i\cap s'_j = \emptyset
    }
    {\Gamma \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\end{gather*}
The first rule states that a paired register is of the product type and its components must be disjoint.
To lift a plain Dirac notation into the labelled version (line 2), we enforce that the term and register share the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
The third line provides the typing of labelled tensor product, with a check to ensure that the component subsystems are disjoint from each other.

@@

The procedure to eliminate labels is sound and complete, in the sense
that expressions in Labelled Dirac notation are equivalent iff their
translations to plain Dirac notation are equivalent.
\begin{theorem}[Label Elimination]
  \label{thm: normalization}
  Assume $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 :
  \cD(s,s')$ and no variables of $\DType(\cdot,\cdot)$ appear in
  $D_1,D_2$.  Let $e_1 = e_2$ be obtained by above normalization
  procedure on $D_1 = D_2$.  Then $e_1=e_2$ is an equation in plain
  Dirac notation and $\Gamma\vDash D_1 = D_2$ if and only if
  $\Gamma\vDash e_1 = e_2$.
\end{theorem}
The idea of the proof is as follows: first, we define a set of proof
rules to rewrite every labelled Dirac notation into the form of
Eqn. (1). We prove that each rule is sound w.r.t.\, semantics, and
that every labelled Dirac expression can be rewritten to an expression
of the form of Eqn. (1); the proof is by induction on the structure of
the expression. Next, we show that reordering of labelled basis
preserves the type and semantics and thus yield expressions of the
form of Eqn. (4), as desired. Details appear in
Appendix~\ref{app:label_elim}
%% \begin{hproof}
%%   \lz{
%%   Checking the soundness of every rule is routine. 
%%   For relative completeness, it is sufficient to show that the normalization steps can succeed for every labelled Dirac equality. To this end, we give 
%%   concrete steps to  by induction on the structure of the notation. Further noticing that reordering of labelled basis preserves the type and semantics and thus yield Eqn. (4). The detailed proof can be found in Appendix \ref{sec: rewriting rules}.
%%   }

### 2025-04-09 05tool_eval.tex
.

\section{Evaluation}\label{sec:eval}
We evaluate D-Hammer on several example sets, and make a comparison with the previous tool DiracDec~\cite{diracdec}.
The experiments are carried out using a MacBook Pro with M3 Max chip. Results are summarized as follows.

@@

The results show significant performance improvements. The enhanced
efficiency primarily results from algorithmic improvements rather than
the shift to a \CC implementation. This is supported by the fact that
efficiency gains are more significant for larger examples, as can be
observed in \Cref{fig: CoqQ plot}.

\subsubsection{Textbook (QCQI)}
As a warm-up, we consider 18 examples from Nielsen and Chuang's
classic texbook~\cite{nielsen2010quantum}. All examples can be encoded
in DiracDec and D-Hammer and are solved very very efficiently.

>>>>

\section{Evaluation}\label{sec:eval}
We evaluate D-Hammer on several example sets, and make a comparison with the previous tool DiracDec~\cite{diracdec}.
The experiments are carried out using a MacBook Pro with M3 Max chip. Results are summarized as follows, which indicates
significant performance improvements.

@@

\subsubsection{Textbook (QCQI)}
As a warm-up, we consider 18 examples from Nielsen and Chuang's
classic texbook~\cite{nielsen2010quantum}. All examples can be encoded
in DiracDec and D-Hammer and are solved very efficiently.
